Tree establishment in models of forest dynamics: a key priority for further research

### 2023-03-30 manuscript/manuscriptRW.Rmd
changes added during the meeting with Harald

There are several factors that lead to mortality during the
establishment phase, such as competition [@casper1997] and multiple
abiotic factors [@schmid2021; @cunningham2006]. As mentioned above, it
is reasonable to expect that models that overestimate ingrowth will
already have a particularly high mortality between the 7 and 10 cm
thresholds. However, this was not consistently evident from the
simulations. This implies that these models must have the excess of
mortality in later stages, if we assume that all models would be able to
capture the structure and composition of adult stands along the EuFoRIa
gradient, which however was not directly tested here.

@@

The EuFoRIa data as used here are unique and well suited to better
understand tree establishment. However, three aspects of these data may
also represent considerable limitations. First, we made a comparison of
tree ingrowth in an equilibrium state, but we cannot assess how close
the forests included in the EuFoRIa dataset were to such a state. The
data were collected in forest reserves where no management has taken
place for long periods of time. This makes our assumption of an
equilibrium between forest properties and environmental drivers more
reasonable than it might appear at first sight. However, the equilibrium
assumption may explain why simulated ingrowth levels were higher than
the observed ones [@brzeziecki2021]. Second, the data were collected for
ingrowth above a 7 cm threshold, thus limiting the assessment of tree
establishment to a holistic viewpoint. This constituted a hard limit
based on which we can understand only some aspects of tree
establishment, which in its entirety often comprises a rather long
period since seed production [@price2001]. In reality, many
environmental constraints are acting on young trees [@käber2021] that we
were unable to assess. Yet, ingrowth data with lower thresholds are
simply not available in a harmonized manner across large environmental
gradients. Third, the empirical data were collected from rather small
plots, while we sampled simulated ingrowth levels from 1 ha areas, which
may lead to an incorrect representation of space. Even though this
strategy is not ideal, it represents a common challenge when harmonizing
diverse data sources originating from varied sampling strategies
[@portier2022]. It would have been extremely challenging for such a
variety of models to follow a protocol where the spatial sampling size
was different at each of the 200 sites, and it would have introduced
additional uncertainty in the results. We must not ignore that the
observed data represent just a snapshot of an extremely complex and
stochastic reality. This, may at least parlty explain e.g. the lack of
clear patterns in the simulation results along the gradients of climate
water balance and degree-days.

@@

This paper arises from a workshop held in Davos (Switzerland) in June
2022, co-led by Olalla Díaz-Yñaez, Yannek Käber and Harald Bugmann.
Funding for the workshop was provided by COST Action CA19139 PROCLIAS
(PROcess-based models for CLimate Impact Attribution across Sectors),
supported by COST (European Cooperation in Science and
Technology; [https://www.cost.eu](https://www.cost.eu/)) and ETH Zurich
(Research Grant ETH-35 18-1). ODY was funded by the Swiss National
Science Foundation project . KB received support from the European
Research Council under the European Union's Horizon 2020 research and
innovation program (Grant Agreement 101001905). JB acknowledges funding
from the long-term research development project no. RVO 67985939 from
the Czech Academy of Sciences. MM acknowledges financial support from
I-Maestro (Innovative forest management strategies for a resilient
bioeconomy under climate change and disturbances, grant no. 22035418,
2019--2022) funded by the ERA-NET Cofund ForestValue. KM received
funding from the Project No. CZ.02.1.01/0.0/0.0/16_019/0000803 financed
by OP RDE, grant no. ITMS2014+313011W580 supported by the Integrated
Infrastructure Operational Programme funded by the ERDF

>>>>

There are several factors that lead to mortality during the
establishment phase, such as competition [@casper1997] and multiple
abiotic factors [@schmid2021; @cunningham2006]. As mentioned above, it
is reasonable to expect that models that overestimate ingrowth may have
a particularly high mortality between the 7 and 10 cm thresholds.
However, this was not consistently evident from the simulations. This
implies that these models must have an excess mortality in later stages,
if we assume that all models were able to capture the structure and
composition of the adult stands along the EuFoRIa gradient; this however
was not tested here.

@@

The EuFoRIa data as used here are unique and well suited to better
understand tree establishment. However, three aspects of these data may
also represent considerable limitations. First, we made a comparison of
tree ingrowth in an equilibrium state, but we cannot assess how close
the forests included in the EuFoRIa dataset were to such a state. The
data were collected in forest reserves where no management has taken
place for long periods of time. This makes our assumption of an
equilibrium between forest properties and environmental drivers more
reasonable than it might appear at first sight. In an analysis of
primeval forests at demographic equilibrium, @brzeziecki2021, they found
higher ingrowth rates than the ones observed in the Euforia data. Thus,
the overestimate of ingrowth rates by the models may not be so
problematic !!! Second, the data were collected for ingrowth above a 7
cm threshold, thus limiting the assessment of tree establishment to a
holistic viewpoint. This constituted a hard limit based on which we can
understand only some aspects of tree establishment, which in its
entirety often comprises a rather long period since seed production
[@price2001]. In reality, many environmental constraints are acting on
young trees [@käber2021] that we were unable to assess. Yet, ingrowth
data with lower thresholds are simply not available in a harmonized
manner across large environmental gradients. Third, the empirical data
were collected from rather small plots, while we sampled simulated
ingrowth levels from 1 ha areas, which may lead to an incorrect
representation of space. Even though this strategy is not ideal, it
represents a common challenge when harmonizing diverse data sources
originating from varied sampling strategies [@portier2022]. It would
have been extremely challenging for such a variety of models to follow a
protocol where the spatial sampling size was different at each of the
200 sites, and it would have introduced additional uncertainty in the
results. We must not ignore that the observed data represent just a
snapshot of an extremely complex and stochastic reality. This, may at
least parlty explain e.g. the lack of clear patterns in the simulation
results along the gradients of climate water balance and degree-days.

@@

This paper arises from a workshop held in Davos (Switzerland) in June
2022, co-led by Olalla Díaz-Yáñez, Yannek Käber and Harald Bugmann.
Funding for the workshop was provided by COST Action CA19139 PROCLIAS
(PROcess-based models for CLimate Impact Attribution across Sectors),
supported by COST (European Cooperation in Science and
Technology; [https://www.cost.eu](https://www.cost.eu/)) and ETH Zurich
(Research Grant ETH-35 18-1). ODY was funded by the Swiss National
Science Foundation project XXXX . KB received support from the European
Research Council under the European Union's Horizon 2020 research and
innovation program (Grant Agreement 101001905). JB acknowledges funding
from the long-term research development project no. RVO 67985939 from
the Czech Academy of Sciences. MM acknowledges financial support from
I-Maestro (Innovative forest management strategies for a resilient
bioeconomy under climate change and disturbances, grant no. 22035418,
2019--2022) funded by the ERA-NET Cofund ForestValue. KM received
funding from the Project No. CZ.02.1.01/0.0/0.0/16_019/0000803 financed
by OP RDE, grant no. ITMS2014+313011W580 supported by the Integrated
Infrastructure Operational Programme funded by the ERDF.
